---
title: "MSstatsSampleSize : A package for optimal design of high-dimensional MS-based proteomics experiment"
author: "Ting Huang (<thuang0703@gmail.com>), Meena Choi (<mnchoi67@gmail.com>), Sumedh Sankhe(<sumedh.sankhe@gmail.com>), Olga Vitek(<o.vitek@northeastern.edu>)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MSstatsSampleSize User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(MSstatsSampleSize)
```


This vignette introduces all the functionalities and summarizes their options in
MSstatsSampleSize. MSstatsSampleSize requires protein abundances quantified in 
mass spectrometry runs as a matrix (columns for biological replicates (samples) 
and rows for proteins) and annotation including Run (Ms runs), biological 
replicates (samples) and their condition (such as a disease and time points).
MSstatsSampleSize includes the following functionalities:

1. `estimateVar`: estimates the variance across biological replicates and MS run
for each protein.
2. `simulateDataset`: simulates data with the given number(s) of biological 
replicates based on the variance estimation. 
3. `designSampleSizeClassification` : fit the classification model (five 
classifier are provided) on each simulation and reports the mean predictive 
accuracy of the classifier and mean protein importance over multiple iterations 
of the simulation. The sample size per condition, which generates the largest 
predictive accuracy, is estimated, while varying the number of biological 
replicates to simulate. Also, the proteins, which can classify the conditions 
best, are reported. The reported sample size per condition can be used to design
future experiments.

In addition, MSstatsTMT includes the following visualization plots for sample 
size estimation: 

1. `meanSDplot`: draw the plot for the mean protein abundance vs standard 
deviation in each condition for the input preliminary dataset. It can exhibit 
the quality of input data.
2. `designSampleSizePCAplot`: make PCA plots for the simulated data with certain
sample size.
3. `designSampleSizeHypothesisTestingPlot`: visualize sample size calculation in
hypothesis testing, which estimates the minimal required number of replicates 
under different expected fold changes.
4. `designSampleSizeClassificationPlots`: visualize sample size calculation in 
classification, including predictive accuracy plot and protein importance plot 
for different sample sizes. 


## 1. Estimate mean protein abundance and variance per condition

### 1.1 estimateVar()

The function fits intensity-based linear model on the preliminary data. This function outputs variance components and mean abundance for each protein.

#### Arguments

* `data` : Data matrix with protein abundance. Rows are proteins and columns are
Biological replicates or samples.
* `annotation` : Group information for samples in data. `Run` for MS run, 
`BioReplicate` for biological subject ID and `Condition` for group information 
are required. `Run` information should be the same with the column of `data`. 
Multiple `Run` may come from same `BioReplicate`.
`log2Trans` : Logical value, if `TRUE` input `data` is log-transformed with base
2. Defaults to `FALSE`

#### Example
```{r}
data("OV_SRM_train")
data("OV_SRM_train_annotation")

head(OV_SRM_train)[,1:5]
head(OV_SRM_train_annotation)

# estimate the mean protein abundance and variance in each condition
variance_estimation <- estimateVar(data = OV_SRM_train, 
                                   annotation = OV_SRM_train_annotation,
                                   log2Trans = FALSE)

# the mean protein abundance in each condition
head(variance_estimation$mu)

# the standard deviation in each condition
head(variance_estimation$sigma)

# the mean protein abundance across all the conditions
head(variance_estimation$promean)

# the standard deviation across all the conditions
head(variance_estimation$prosd)
```


### 1.2 meanSDplot()

This function draws the plot for the mean protein abundance (X-axis) vs standard
deviation (Y-axis) in each condition. The `lowess` function is used to fit the 
LOWESS smoother between mean protein abundance and standard deviation (square 
root of variance). This function generates the mean-SD plot or a pdf file based
on the selected inputs.


#### Arguments

* `data` : A list with mean protein abundance matrix and standard deviation 
matrix. It should be the output of `estimateVar` function.
* `x.axis.size` : Size of x-axis labeling in Mean-SD Plot. Default is 10.
* `y.axis.size` : Size of y-axis labels. Default is 10.
* `smoother_size` : Size of lowess smoother. Default is 1.
* `width` : Width of the saved pdf file. Default is 4.
* `height` : Height of the saved pdf file. Default is 4.
* `xlimUp` : The upper limit of x-axis for mean-SD plot. Default is 30.
* `ylimUp` : The upper limit of y-axis for mean-SD plot. Default is 3.
* `address` : The name of folder that will store the results. Default folder is 
the current working directory. The other assigned folder has to be existed under
the current working directory. An output pdf file is automatically created with 
the default name of `MeanSDPlot.pdf`. The command address can help to specify 
where to store the file as well as how to modify the beginning of the file name.
If address=FALSE, plot will be not saved as pdf file but showed in window.

#### Example
```{r, eval=FALSE}
#  output a pdf file with mean-SD plot
meanSDplot(variance_estimation)
```



## 2. Simulates data with the given numbers of biological replicates and proteins
based on the variance estimation

### 2.1 simulateDataset()

This function simulate datasets with the given numbers of biological replicates 
and proteins based on the preliminary dataset (input for this function). The 
function fits intensity-based linear model on the input data `data` in order to 
get variance and mean abundance, using `estimateVar` function. Then it uses 
variance components and mean abundance to simulate new training data with the 
given sample size and protein number. It outputs the number of simulated proteins, 
a vector with the number of simulated samples in each condition, the list of 
simulated training datasets, the input dataset and the (simulated) validation 
dataset.


#### Arguments

* `data` : Protein abundance data matrix. Rows are proteins and columns are biological replicates (samples).
* `annotation` : Group information for samples in data.m`Run' for MS run, `BioReplicate' for biological subject ID and `Condition' for group information are required. `Run' information should be the same with the column of `data'. Multiple `Run' may come from same `BioReplicate'.
* `log2Trans` : Default is FALSE. If TRUE, the input `data' is log-transformed with base 2.
* `num_simulations` : Number of times to repeat simulation experiments (Number of simulated datasets). Default is 10.
* `samples_per_group` : Number of samples per group to simulate. Default is 50.
* `protein_rank` : The standard to rank the proteins in the input `data'. It can be 1) "mean" of protein abundances over all the samples or 2) "sd" (standard deviation) of protein abundances over all the samples or 3) the "combined" of mean abundance and standard deviation.
The proteins in the input `data' are ranked based on `protein_rank' and the user can select a subset of proteins to simulate.
* `protein_select` : select proteins with "low" or "high" mean abundance or standard deviation (variance) or their combination to simulate. If `protein_order = "mean"' or protein_order = "sd"', `protein_select' should be "low" or "high". Default is "high", indicating high abundance or standard deviation proteins are selected to simulate.
If `protein_order = "combined"', `protein_select' has two elements. The first element corrresponds to the mean abundance. The second element corrresponds to the standard deviation (variance). Default is c("high", "low") (select proteins with high abundance and low variance).
* `protein_quantile_cutoff` : Quantile cutoff(s) for selecting protiens to simulate. For example, when `protein_rank="mean"', and `protein_select="high"', `protein_quantile_cutoff=0.1' Proteins are ranked based on their mean abundance across all the samples. Then, the top 10% highest abundant proteins are selected to simulate. Default is 0.0, which means that all the proteins are used. If `protein_rank = "combined"', `protein_quantile_cutoff'` has two cutoffs. The first element corrresponds to the cutoff for mean abundance. The second element corrresponds to the cutoff for the standard deviation (variance). Default is c(0.0, 1.0), which means that all the proteins will be used.
* `expected_FC` : Expected fold change of proteins. The first option (Default) is "data", indicating the fold changes are directly estimated from the input `data'. The second option is a vector with predefined fold changes of listed proteins. The vector names must match with the unique information of Condition in `annotation'. One group must be selected as a baseline and has fold change 1 in the vector. The user should provide list_diff_proteins, which users expect to have the fold changes greater than 1. Other proteins that are not available in `list_diff_proteins' will be expected to have fold change = 1
* `list_diff_proteins` : Vector of proteins names which are set to have fold changes greater than 1 between conditions. If user selected `expected_FC= "data" ', this should be NULL.
* `simulate_validation` : Default is FALSE. If TRUE, simulate the validation set; otherwise, the input `data' will be used as the validation set.
* `valid_samples_per_group` : Number of validation samples per group to simulate. This option works only when user selects `simulate_validation=TRUE'. Default is 50.


#### Example

```{r message = FALSE , warning = FALSE}
# expected_FC = "data": fold change estimated from OV_SRM_train
# select_simulated_proteins = "proportion": select the simulated proteins based 
#on the proportion of total proteins
# simulate_valid = FALSE: use input OV_SRM_train as validation set
simulated_datasets <- simulateDataset(data = OV_SRM_train,
                                      annotation = OV_SRM_train_annotation,
                                      log2Trans = FALSE,
                                      num_simulations = 10, # simulate 10 times
                                      samples_per_group = 50, # 50 samples per condition
                                      protein_rank = "mean",
                                      protein_select = "high",
                                      protein_quantile_cutoff = 0.0,
                                      expected_FC = "data",
                                      list_diff_proteins =  NULL,
                                      simulate_validation = FALSE,
                                       valid_samples_per_group = 50)

```

Explore the output from `simulateDataset` function
```{r}
# the number of simulated proteins
simulated_datasets$num_proteins

# a vector with the number of simulated samples in each condition
simulated_datasets$num_samples

# the list of simulated protein abundance matrices
# Each element of the list represents one simulation
head(simulated_datasets$simulation_train_Xs[[1]]) # first simulation

# the list of simulated condition vectors
# Each element of the list represents one simulation
head(simulated_datasets$simulation_train_Ys[[1]]) # first simulation
```

User can also specify the expected fold change of proteins they consider to be 
deferentially abundant between conditions.
```{r message = FALSE , warning = FALSE}
# expected_FC = expected_FC: user defined fold change
unique(OV_SRM_train_annotation$Condition)
expected_FC <- c(1, 1.5)
names(expected_FC) <- c("control", "ovarian cancer")
set.seed(1212)
# Here I randomly select some proteins as differential to show how the function works
# The user should prepare this list of differential proteins by themselves
diff_proteins <- sample(rownames(OV_SRM_train), 20)
simualted_datasets_predefined_FC <- simulateDataset(data = OV_SRM_train,
                                      annotation = OV_SRM_train_annotation,
                                      log2Trans = FALSE,
                                      num_simulations = 10, # simulate 10 times
                                      samples_per_group = 50, # 50 samples per condition
                                      protein_rank = "mean",
                                      protein_select = "high",
                                      protein_quantile_cutoff = 0.0,
                                      expected_FC = expected_FC,
                                      list_diff_proteins =  diff_proteins,
                                      simulate_validation = FALSE,
                                       valid_samples_per_group = 50)


```



## 3. Sample size estimation for classification

### 3.1. designSampleSizeClassification()

This function fits the classification model, in order to classify the subjects 
in the simulated training datasets (in the output of `simulatedDataset`). Then 
the fitted model is validated by the (simulated) validation set. Two performance
are reported :
(1) the mean predictive accuracy : The function trains classifier on each 
simulated training dataset and reports the predictive accuracy of the trained 
classifier on the validation data (output of `SimulateDataset` function). Then 
these predictive accuracies are averaged over all the simulation.
(2) the mean protein importance : It represents the importance of a protein in 
separating different groups. It is estimated on each simulated training dataset 
using function `varImp` from package caret. Please refer to the help file of 
`varImp` about
how each classifier calculates the protein importance. Then these importance 
values for each protein are averaged over all the simulation.

The list of classification models trained on each simulated dataset, the 
predictive accuracy on the validation set predicted by the corresponding 
classification model and the importance value for all the proteins estimated by 
the corresponding classification model are also reported.

### Arguments

* `simulations` : A list of simulated datasets It should be the name of the 
output of `SimulateDataset` function.
* `classifier` : A string specifying which classfier to use. This function uses 
function `train` from package caret. The options are 1) rf (random forest 
classifier, default option). 2) nnet (neural network), 3) svmLinear (support 
vector machines with linear kernel), 4) logreg(logistic regression), and 5) 
naive_bayes (naive_bayes).
* `parallel` : Default is FALSE. If TRUE, parallel computation is performed.


### Example

```{r message = FALSE, warning = FALSE}
classification_results <- designSampleSizeClassification(
    simulations = simulated_datasets, 
    parallel = FALSE)

```

Explore the output of `designSampleSizeClassification`
```{r message = FALSE, warning = FALSE}
# the number of simulated proteins
classification_results$num_proteins
# a vector with the number of simulated samples in each condition
classification_results$num_samples
# the mean predictive accuracy over all the simulated datasets,
# which have same 'num_proteins' and 'num_samples'
classification_results$mean_predictive_accuracy
# the mean protein importance vector over all the simulated datasets,
# the length of which is 'num_proteins'.
head(classification_results$mean_feature_importance)
```

In order to speed up the running time, the package also provides parallel 
computation for `designSampleSizeClassification` function.
```{r message = FALSE, warning = FALSE}
## try parallel computation to speed up
## The parallel computation may cause error while allocating the core resource
## Then the users can try the above function without parallel computation
classification_results_parallel <- designSampleSizeClassification(
    simulations = simulated_datasets, 
    parallel = TRUE)
```


### 3.2 designSampleSizeClassificationPlots()

This function visualizes for sample size calculation in classification. Mean 
predictive accuracy and mean protein importance under each sample size is from 
the input `data`, which is the output from function `designSampleSizeClassification`.
To illustrate the mean predictive accuracy and protein importance under different
sample sizes, it generates two types of plots in pdf files as output : 

(1) The predictive accuracy plot shows the mean predictive accuracy under 
different sample sizes. The X-axis represents different sample sizes and y-axis 
represents the mean predictive accuracy.

(2) The protein importance plot includes multiple subplots. The number of 
subplots is equal to `list_samples_per_group`. Each subplot shows the top 
`num_important_proteins_show` most important proteins under each sample size. 
The Y-axis of each subplot is the protein name and X-axis is the mean protein 
importance under the sample size.

(3) A numeric value which is the estimated optimal sample size per group for the
input dataset for classification problem.

While varying the number of biological replicates to simulate, the sample size 
per condition which generates the largest predictive accuracy can be found from 
the predictive accuracy plot, The optimal sample size per condition can be used 
to design future experiments. Also, the proteins, which can classify the 
conditions best, are reported by the protein importance plot. 

#### Arguments

* `data` : A list of outputs from function `designSampleSizeClassification`. 
Each element represents the results under a specific sample size. The input 
should include at least two simulation results with different sample sizes.
* `list_samples_per_group` :  A vector includes the different sample sizes 
simulated. This is required. The number of simulation in the input `data` should
be equal to the length of list_samples_per_group
* `optimal_threshold`: The maximal cutoff for deciding the optimal sample size. 
Default is 0.0001. Large cutoff can lead to smaller optimal sample size
whereas small cutoff produces large optimal sample size.
* `num_important_proteins_show` : The number of proteins to show in protein 
importance plot.
* `protein_importance_plot` : TRUE(default) draws protein importance plot.
* `predictive_accuracy_plot` : TRUE(default) draws predictive accuracy plot.
* `x.axis.size` : Size of x-axis labeling in predictive accuracy plot and protein
importance plot. Default is 10.
* `y.axis.size` : Size of y-axis labels in predictive accuracy plot and protein 
importance plot. Default is 10.
* `predictive_accuracy_plot_width` : Width of the saved pdf file for predictive 
accuracy plot. Default is 4.
* `predictive_accuracy_plot_height` : Height of the saved pdf file for 
predictive accuracy plot. Default is 4.
* `protein_importance_plot_width` : Width of the saved pdf file for protein 
importance plot. Default is 3.
* `protein_importance_plot_height` : Height of the saved pdf file for protein 
importance plot. Default is 3.
* `ylimUp_predictive_accuracy` : The upper limit of y-axis for predictive 
accuracy plot. Default is 1. The range should be 0 to 1.
* `ylimDown_predictive_accuracy` : The lower limit of y-axis for predictive 
accuracy plot. Default is 0.0. The range should be 0 to 1.
* `address` : The name of folder that will store the results. Default folder is 
the current working directory. The other assigned folder has to be existed under
the current working directory. An output pdf file is automatically created with 
the default name of `PredictiveAccuracyPlot.pdf` and `ProteinImportancePlot.pdf`.
The command address can help to specify where to store the file as well as how 
to modify the beginning of the file name. If address=FALSE, plot will be not 
saved as pdf file but showed in window.


#### Example

```{r, eval=FALSE}
#### sample size classification ####
# simulate different sample sizes
# 1) 10 biological replicates per group
# 1) 25 biological replicas per group
# 2) 50 biological replicas per group
# 3) 100 biological replicas per group
# 4) 200 biological replicas per group
list_samples_per_group <- c(10, 25, 50, 100, 200)

# save the simulation results under each sample size
multiple_sample_sizes <- list()

for(i in seq_along(list_samples_per_group)){
    # run simulation for each sample size
    simulated_datasets <- simulateDataset(data = OV_SRM_train,
                                          annotation = OV_SRM_train_annotation,
                                          log2Trans = FALSE,
                                          num_simulations = 10, # simulate 10 times
                                          samples_per_group = list_samples_per_group[i],
                                          protein_rank = "mean",
                                          protein_select = "high",
                                          protein_quantile_cutoff = 0.0,
                                          expected_FC = "data",
                                          list_diff_proteins =  NULL,
                                          simulate_valid = FALSE,
                                          valid_samples_per_group = 50)

    # run classification performance estimation for each sample size
    res <- designSampleSizeClassification(simulations = simualted_datasets,
                                          parallel = TRUE)
    
    # save results
    multiple_sample_sizes[[i]] <- res
}

## make the plots
designSampleSizeClassificationPlots(multiple_sample_sizes,
                                    list_samples_per_group,
                                    ylimUp_predictive_accuracy = 0.8,
                                    ylimDown_predictive_accuracy = 0.6)
```

## 4. Sample size estimation for hypothesis testing

### 4.1. designSampleSizeHypothesisTestingPlot()

The function fits intensity-based linear model on the input `data`. Then it uses
the fitted models and the fold changes estimated from the models to calculate 
sample size for hypothesis testing through `designSampleSize` function from 
MSstats package. It outputs a table with the minimal number of biological 
replicates per condition to acquire the expected FDR and power under different 
fold changes, and a PDF file with the sample size plot.

#### Arguments
* `data` : Protein abundance data matrix.
Rows are proteins and columns are biological replicates (samples).
* `annotation` : Group information for samples in data. `Run` for MS run, `BioReplicate` for biological subject ID and `Condition` for group information are required. `Run` information should be the same with the column of `data`. Multiple `Run` may come from same `BioReplicate`.
* `log2Trans` : Default is FALSE. If TRUE, the input `data' is log-transformed with base 2.
* `desired_FC` : the range of a desired fold change. The first option (Default) is "data", indicating the range of the desired fold change is directly estimated from the input `data`, which are the minimal fold change and the maximal fold change in the input `data'. The second option is a vector which includes the lower and upper values of the desired fold change (For example, c(1.25,1.75)).
* `protein_rank` : The standard to rank the proteins in the input `data'. It can be 1) "mean" of protein abundances over all the samples or 2) "sd" (standard deviation) of protein abundances over all the samples or 3) the "combined" of mean abundance and standard deviation. The proteins in the input `data' are ranked based on `protein_rank'
and the user can select a subset of proteins for hypothesis testing and sample size calculation.
* `protein_select` : select proteins with "low" or "high" mean abundance or standard deviation (variance) or their combination for hypothesis testing and sample size calculation. The variance (and the range of desired fold change if desiredFC = "data") will be estimated from the selected proteins. If `protein_order = "mean"' or protein_order = "sd"', `protein_select' should be "low" or "high". Default is "high", indicating high abundance or standard deviation proteins are selected. If `protein_order = "combined"', `protein_select' has two elements. The first element corrresponds to the mean abundance. The second element corrresponds to the standard deviation (variance). Default is c("high", "low") (select proteins with high abundance and low variance).
* `protein_quantile_cutoff` : Quantile cutoff(s) for selecting protiens for hypothesis testing and sample size calculation. For example, when `protein_rank="mean"', and
`protein_select="high"', `protein_quantile_cutoff=0.1' Proteins are ranked based on their mean abundance across all the samples. Then, the top 10% highest abundant proteins are selected. Default is 0.0, which means that all the proteins are used. If `protein_rank = "combined"', `protein_quantile_cutoff'` has two cutoffs. The first element corrresponds to the cutoff for mean abundance. The second element corrresponds to the cutoff for the standard deviation (variance). Default is c(0.0, 1.0), which means that all the proteins will be used.
* `FDR` : a pre-specified false discovery ratio (FDR) to control the overall false positive. Default is 0.05.
* `power` : a pre-specified statistical power which defined as the probability of detecting a true fold change. You should input the average of power you expect. Default is 0.9.
* `width` : Width of the saved pdf file. Default is 5.
* `height` : Height of the saved pdf file. Default is 5.
* `address` : The name of folder that will store the results. Default folder is 
the current working directory. The other assigned folder has to be existed under
the current working directory. An output pdf file is automatically created with 
the default name of `HypothesisTestingSampleSizePlot.pdf`. The command address 
can help to specify where to store the file as well as how to modify the beginning
of the file name.
If address=FALSE, plot will be not saved as pdf file but showed in window.

#### Example
```{r}
#  output a pdf file with sample size calculation plot for hypothesis testing
#  also return a table which summaries the plot
HT_res <- designSampleSizeHypothesisTestingPlot(data = OV_SRM_train,
                                                annotation= OV_SRM_train_annotation,
                                                desired_FC = "data",
                                                protein_rank = "mean",
                                                protein_select = "high",
                                                protein_quantile_cutoff = 0.0,
                                                FDR=0.05,
                                                power=0.9)

# data frame with columns desiredFC, numSample, FDR, power and CV
head(HT_res)
```

## 5. Visualization for simulated datasets

### 5.1 designSampleSizePCAplot()

This function draws PCA plots for the preliminary dataset and each simulated 
dataset in `simulations` (input for this function). It outputs a pdf file where 
the number of page is equal to the number of simulations plus 1. The first page 
represents a PCA plot for the input `data` (`OV_SRM_train`). Each of the following
pages presents a PCA plot under one simulation. X-axis of PCA plot is the first
component and y-axis is the second component. This function can be used to 
validate whether the simulated dataset looks consistent with the input 
preliminary dataset.

#### Arguments
* `simulations` : A list of simulated datasets. It should be the output of 
`simulateDataset` function.
* `which.PCA` : Select one PCA plot to show. It can be "all", "allonly", or 
"simulationX". X should be index of simulation, such as "simulation1" or 
"simulation5". Default is "all", which generates all the plots. "allonly" 
generates the PCA plot for the whole input dataset. "simulationX" generates 
the PCA plot for a specific simulated dataset (given by index).
* `x.axis.size` : Size of x-axis labeling in PCA Plot. Default is 10.
* `y.axis.size` : Size of y-axis labels. Default is 10.
* `dot.size` : Size of dots in PCA plot. Default is 3.
* `legend.size` : Size of legend above Profile plot. Default is 7.
* `width` : Width of the saved pdf file. Default is 6.
* `height` : Height of the saved pdf file. Default is 5.
* `address` : The name of folder that will store the results. Default folder is 
the current working directory. The other assigned folder has to be existed under
the current working directory. An output pdf file is automatically created with 
the default name of `PCAPlot.pdf`. The command address can help to specify where
to store the file as well as how to modify the beginning of the file name. 
If address=FALSE, plot will be not saved as pdf file but showed in window.

#### Example
```{r, eval=FALSE}
#  output a pdf file with 11 PCA plots
designSampleSizePCAplot(simulated_datasets)
```
